Eco-evolutionary dynamics of finite populations from first principles

### 2023-02-08 ind_files/06_unification_discussion.tex
Fix minor typo in stochastic RM/Price

&+ \frac{1}{\sqrt{K}}\underbrace{\left[\left(A^{+}_{i}\right)^{1/2}dB^{(i)}_t - p_i\sum\limits_{j=1}^{m}\left(A^{+}_{j}\right)^{1/2}dB^{(j)}_t\right]}_{\substack{\text{Non-directional finite population effects}\\\text{due to stochastic fluctuations}}}

@@

&+ \frac{1}{\sqrt{K}}\left(\sum\limits_{i=1}^{m}f_i\sqrt{A_i^+}dB_{t}^{(i)} - \overline{f}\sum\limits_{i=1}^{m}\sqrt{A_i^+}dB_{t}^{(i)}\right)

@@

where all covariances are understood in the statistical sense (Note that since $w_i$, $\tau_i$, $\overline{w}$, and $\overline{\tau}$ are stochastic processes depending on $\mathbf{p}$, the terms $ \textrm{Cov}(w,f)$ and $\textrm{Cov}(\tau,f)$ are themselves stochastic processes). Taking $K \to \infty$ in equation \eqref{nD_stochastic_Price} recovers the standard Price equation as the infinite population limit. In some sense, \eqref{nD_stochastic_RM} and \eqref{nD_stochastic_Price} are therefore `fundamental' equations for the evolution of finite populations, with the replicator-mutator and Price equations as the infinite population limits respectively. In both equations, the first term represents the direct effects of forces captured in classic deterministic models - the selection and mutation are partitioned into two terms in \eqref{nD_stochastic_Price}, but since mutation acts symmetrically in both the first and second term of \eqref{nD_stochastic_RM}, this has no major bearing on our interpretations. As is evident in both \eqref{nD_stochastic_RM} and \eqref{nD_stochastic_Price}, finite populations experience a new directional force dependent on $\tau_i(\mathbf{x}) = b^{\textrm{(ind)}}_{i}(\mathbf{x}) + b^{\textrm{(ind)}}_{i}(\mathbf{x})$, the per-capita turnover rate of type $i$, that cannot be captured in infinite population models. Remarkably, this term acts in a way that is mathematically exactly analogously to the classical evolutionary force of selection and mutation as captured by the first term in \eqref{nD_stochastic_RM}, but in the opposite direction - A higher relative $\tau_i$ leads to a decrease in frequency (Notice the minus sign before the second term in \eqref{nD_stochastic_RM}). Biologically, this term describes a biasing effect due to differential turnover rates and can intuitively be understood as being similar to the gambler's ruin through the following reasoning: If a type $i$ has a higher $\tau_i$, it experiences greater turnover due to a generally higher birth and death rate and thus experience more births and deaths in a given time interval than an otherwise equivalent species with a lower $\tau_i$. More events mean greater demographic stochasticity, and types with a higher $\tau_i$ thus tend to be eliminated by a stochastic analog of selection because they experience more chance events (births and deaths) in a given time period. This effect is less visible if the total population size is higher because larger populations generally experience less stochasticity, which is reflected in the $1/N_K$ factor in this term. This stochastic analog of selection for reduced turnover rates, captured by the second term of equation \eqref{nD_stochastic_RM}, is the force responsible for the `reversal of the direction of deterministic selection' induced by demographic noise in previous studies \citep{houchmandzadeh_selection_2012, houchmandzadeh_fluctuation_2015, constable_demographic_2016, mcleod_social_2019}. Note that types that tend to increase the \emph{total} population size $N_K(t)$ (such as altruists in evolutionary theory and mutualists in ecological communities) will reduce the magnitude of this effect compared to types that do not facilitate such an increase, such as cheaters and highly competitive species, which could explain why this effect preferentially favors the former types in reversing the direction of deterministic selection in finite population models \citep{houchmandzadeh_fluctuation_2015, mcleod_social_2019}. The fact that total population size controls the strength of the second term of \eqref{nD_stochastic_RM} also explains why cooperation is favoured in the early transient period of population growth \citep{melbinger_evolutionary_2010} when simulations are initiated from a small population size - In the early transient period, $N_K(t)$ is small, and the biasing effect of differential turnover rates is stronger, thus favouring cooperation. More generally, selection for reduced turnover rate could explain why cooperation often persists for a long time in finite population IbMs (and the real world) despite infinite population models predicting their extinction. The fact that the entire term is multiplied by $(KN_K(t))^{-1}$ suggests that the effect of this force is weak for medium to large populations, which explains why the persistence of cooperators is often only observed in restrictive sounding conditions such as quasi-neutrality, rapid attraction to a slow manifold, or a weak selection + weak mutation limit. In all three of these cases, the first term on the RHS of \eqref{nD_eqn_for_frequencies} becomes identically 0. It therefore no longer contributes to the total dynamics and allows us to see the contributions of the second term.\\

>>>>

&+ \frac{1}{\sqrt{K}N_{K}(t)}\underbrace{\left[\left(A^{+}_{i}\right)^{1/2}dB^{(i)}_t - p_i\sum\limits_{j=1}^{m}\left(A^{+}_{j}\right)^{1/2}dB^{(j)}_t\right]}_{\substack{\text{Non-directional finite population effects}\\\text{due to stochastic fluctuations}}}

@@

&+ \frac{1}{\sqrt{K}N_{K}(t)}\left(\sum\limits_{i=1}^{m}f_i\sqrt{A_i^+}dB_{t}^{(i)} - \overline{f}\sum\limits_{i=1}^{m}\sqrt{A_i^+}dB_{t}^{(i)}\right)

@@

where all covariances are understood in the statistical sense (Note that since $w_i$, $\tau_i$, $\overline{w}$, and $\overline{\tau}$ are stochastic processes depending on $\mathbf{p}$, the terms $ \textrm{Cov}(w,f)$ and $\textrm{Cov}(\tau,f)$ are themselves stochastic processes). Taking $K \to \infty$ in equation \eqref{nD_stochastic_Price} recovers the standard Price equation as the infinite population limit. An infinite-dimensional version of \eqref{nD_stochastic_Price} has recently also been shown to hold for one-dimensional quantitative traits when $f = x$, the trait value, under certain assumptions \citep{week_white_2021}. In some sense, \eqref{nD_stochastic_RM} and \eqref{nD_stochastic_Price} are therefore `fundamental' equations for the evolution of finite populations, with the replicator-mutator and Price equations as their respective infinite population limits. In both equations, the first term represents the direct effects of forces captured in classic deterministic models - the selection and mutation are partitioned into two terms in \eqref{nD_stochastic_Price}, but since mutation acts symmetrically in both the first and second term of \eqref{nD_stochastic_RM}, this has no major bearing on our interpretations. As is evident in both \eqref{nD_stochastic_RM} and \eqref{nD_stochastic_Price}, finite populations experience a new directional force dependent on $\tau_i(\mathbf{x}) = b^{\textrm{(ind)}}_{i}(\mathbf{x}) + b^{\textrm{(ind)}}_{i}(\mathbf{x})$, the per-capita turnover rate of type $i$, that cannot be captured in infinite population models. Remarkably, this term acts in a way that is mathematically exactly analogously to the classical evolutionary force of selection and mutation as captured by the first term in \eqref{nD_stochastic_RM}, but in the opposite direction - A higher relative $\tau_i$ leads to a decrease in frequency (Notice the minus sign before the second term in \eqref{nD_stochastic_RM}). Biologically, this term describes a biasing effect due to differential turnover rates and can intuitively be understood as being similar to the gambler's ruin through the following reasoning: If a type $i$ has a higher $\tau_i$, it experiences greater turnover due to a generally higher birth and death rate and thus experience more births and deaths in a given time interval than an otherwise equivalent species with a lower $\tau_i$. More events mean greater demographic stochasticity, and types with a higher $\tau_i$ thus tend to be eliminated by a stochastic analog of selection because they experience more chance events (births and deaths) in a given time period. This effect is less visible if the total population size is higher because larger populations generally experience less stochasticity, which is reflected in the $1/N_K$ factor in this term. This stochastic analog of selection for reduced turnover rates, captured by the second term of equation \eqref{nD_stochastic_RM}, is the force responsible for the `reversal of the direction of deterministic selection' induced by demographic noise in previous studies \citep{houchmandzadeh_selection_2012, houchmandzadeh_fluctuation_2015, constable_demographic_2016, mcleod_social_2019}. Note that types that tend to increase the \emph{total} population size $N_K(t)$ (such as altruists in evolutionary theory and mutualists in ecological communities) will reduce the magnitude of this effect compared to types that do not facilitate such an increase, such as cheaters and highly competitive species, which could explain why this effect preferentially favors the former types in reversing the direction of deterministic selection in finite population models \citep{houchmandzadeh_fluctuation_2015, mcleod_social_2019}. The fact that total population size controls the strength of the second term of \eqref{nD_stochastic_RM} also explains why cooperation is favoured in the early transient period of population growth \citep{melbinger_evolutionary_2010} when simulations are initiated from a small population size - In the early transient period, $N_K(t)$ is small, and the biasing effect of differential turnover rates is stronger, thus favouring cooperation. More generally, selection for reduced turnover rate could explain why cooperation often persists for a long time in finite population IbMs (and the real world) despite infinite population models predicting their extinction. The fact that the entire term is multiplied by $(KN_K(t))^{-1}$ suggests that the effect of this force is weak for medium to large populations, which explains why the persistence of cooperators is often only observed in restrictive sounding conditions such as quasi-neutrality, rapid attraction to a slow manifold, or a weak selection + weak mutation limit. In all three of these cases, the first term on the RHS of \eqref{nD_eqn_for_frequencies} becomes identically 0. It therefore no longer contributes to the total dynamics and allows us to see the contributions of the second term.\\
